---
title: 'Homework #4'
author: "Kaitlyn Harper"
date: "October 5, 2017"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General data reshaping and relationship to aggregation

Activity #1 - make a cheatsheet for the tidy package (gather and spread functions)

Hi there! Welcome to my tidy data cheatsheet. I got most of my information from [this](http://vita.had.co.nz/papers/tidy-data.pdf) paper, written by the one and only Hadley Wickham. He goes into a lot more detail and uses large data sets as examples, so if you find yourself itching for more information, I'd definitely recommend checking it out! 

Okay then, let's get going! 

- make two examples of messy data 
- plot messy data
- fix messy data (make tidy data)
- plot tidy data

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table

Five most common problems of messy data:
• Column headers are values, not variable names. (treatmenta, treatmentb, placebo - with frequency in each one)
• Multiple variables are stored in one column. (maleGene1, femaleGene1, maleGene2, femaleGene2)
• Variables are stored in both rows and columns. (within age column: c(m14, m21, f15, f18)
• Multiple types of observational units are stored in the same table. element: c(tmax, tmin, tmax, tmin) ... unit: c(27, 21, 29, 22)
• A single observational unit is stored in multiple tables. 


```{r}

```

