---
title: 'Homework #2'
author: "Kaitlyn Harper"
date: "September 18, 2017"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bring rectangular data in
```{r load_data, echo=TRUE}

# Load tidyverse + all corresponding packages!
library(tidyverse)

# Load gapminder dataset
library(gapminder)
data("gapminder")

#View data (just to make sure it's there)
head(gapminder)

```

<a href="#top">Back to top</a>

## "Smell test" the data
**Q1: Is it a data.frame, a matrix, a vector, a list?**
**Q2: What’s its class?**
```{r gapminder_class, echo=TRUE}

typeof(gapminder) #you can check the type of data with this
class(gapminder) #this is another easy way to check the class of data

```
The type of object is a `r typeof(gapminder)`, but it has three official classes: *data frame tbl, tbl,* and *data frame*.

**Q3: How many variables/columns?**
**Q4: How many rows/observations?**
```{r variable_count, echo=TRUE}

ncol(gapminder) # You can do this to check the number of columns
nrow(gapminder) # Plus this checks the number of rows

dim(gapminder) # Or just do this! It will tell you both! 

```
There are `r ncol(gapminder)` variables/columns and `r nrow(gapminder)` rows in this data set.

**Q5: Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?**  
Let's take a "glimpse" at the data! :)
```{r other_ways, echo=TRUE}

glimpse(gapminder) # a helpful function for looking at extent/size
str(gapminder) # this is another function for looking at extent/size

```
One way you could get these dimensions (and a bunch of other info about the data set) is by using the glimpse or str functions. These function provide the class of each variable, the number of different values for factor variables, and the first few rows for each variable. This could be helpful when you want to find out more info with fewer clicks, and it can also help us answer the next question! :)

**Q6: What flavor is each variable?**  
To check the "flavor", I looked at the class of each variable. Check out my original [Rmd file](https://github.com/oktokat/STAT545-hw-harper-kaitlyn/blob/master/hw02/hw02.Rmd) to see how I inserted code into this paragraph!! The `r names(gapminder)[1]` and `r names(gapminder)[2]` variables are **`r paste(class(gapminder$country), "s", sep = "")`**; the `r names(gapminder)[3]` and `r names(gapminder)[5]` variables are **`r paste(class(gapminder$year), "s", sep = "")`**; the `r names(gapminder)[4]` and `r names(gapminder)[6]` variables are **`r class(gapminder$lifeExp)`**.

<a href="#top">Back to top</a>

## Explore individual variables  
**Pick at least one categorical variable and at least one quantitative variable to explore.**    
I will explore the country variable and the life expectancy variable.

**Characterize what’s possible, i.e. all possible values or max vs. min … whatever’s appropriate.**
```{r variable_info, echo=TRUE}

# Qualitative variable: country
head(unique(gapminder$country)) # shows the unique values within the country variable (only showing the first six)
length(unique(gapminder$country)) # number of unique values within the country variable
head(table(gapminder$country)) # number of rows for each country (only showing first six)
numCountries = table(gapminder$continent)/(table(gapminder$country)[1]) # number of countries within each continent

# Quantitative variable: life expectancy 
summary(gapminder$lifeExp) # provides a simple statistical summary of the variable
length(unique(round(gapminder$lifeExp, 1))) # provides the number of unique life expectancies for all rows (note: I rounded the numbers to provide a more general statistic)

```
*Qualitative variable: country*  
There are `r length(unique(gapminder$country))` unique countries in this data set. There are 195 countries in the world, so this data set is missing `r 195-length(unique(gapminder$country))` countries. Each country contains `r table(gapminder$country)[1]` rows, or `r table(gapminder$country)[1]` different points in time. There are `r length(unique(gapminder$continent))` continents represented in this data set. There are `r numCountries[1]` countries represented from `r names(numCountries[1])`, `r numCountries[2]` countries represented from the `r names(numCountries[2])`, `r numCountries[3]` countries represented from `r names(numCountries[3])`, `r numCountries[4]` countries represented from `r names(numCountries[4])`, and `r numCountries[5]` countries represented from `r names(numCountries[5])`

*Quantitative variable: life expectancy*  
First, we'll look at the life expectancy variable as a whole. The minimum life expectancy from all countries, all years was `r round(summary(gapminder$lifeExp)[1], 2)` and the maximum life expectancy was `r round(summary(gapminder$lifeExp)[6], 2)`. The mean life expectancy was `r round(summary(gapminder$lifeExp)[4], 2)`. However, this isn't all that helpful to us, because there are multiple countries and multiple time points measured. A bit later, we'll break up some of the countries and time points and get a better look! 

**What’s typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.**

```{r simple_plots, echo=TRUE}

# Plot the life expectancy of each country
gapminder %>%                                                 # pipe in data set
  ggplot(aes(country, lifeExp)) +                             # insert general aesthetics 
  geom_point() +                                              # add points
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  # arrange text vertically
  ylab("life expectancy (years)") +                           # add y-axis label
  ggtitle("Life expectancy of each country from 1952-2007")   # add title

```

Looks like the distribution of life expectancies varies a LOT between countries and also within the same country! Check out the plots below to get a better feel for the data! 

<a href="#top">Back to top</a>

## Explore various plot types  
Here are a few different plots using all the data (no filtering/grouping yet)

### Two continuous variables
```{r scatterplot, echo=TRUE}

# Create an object with the data/aesthetics
a = ggplot(gapminder, aes(x= year, y = lifeExp))

# Plot life expectancy with unadjusted abline
a + geom_point() + 
  labs(title = "A look at life expectancy over the years 1952-2007, unadjusted", y = "life expectancy (years)") + stat_smooth(method="lm", se=FALSE, color = "lightblue", size = 2)

# Use linear regression to adjust for gdp per capita
reg = lm(gapminder$lifeExp ~ gapminder$year + gapminder$gdpPercap)
summary(reg)
coeffs = coefficients(reg)

# Plot life expectancy with abline, adjusted for GDP per capita
a + geom_point() + 
  labs(title = "A look at life expectancy over the years 1952-2007, adjusted for GDP per capita", y = "life expectancy (years)") + geom_abline(intercept = coeffs[1], slope = coeffs[2], color = "plum", size = 2)

```

### One qualitative variable
```{r histogram, echo=TRUE}

# Histogram of the distribution of life expectancies
ggplot(gapminder, aes(lifeExp)) +
  geom_histogram(binwidth = 2) +
  labs(y = "Count", x = "Life expectancy (years)", title= "Distribution of life expectancies")

# Barplot of the number of country observations within each continent
ggplot(gapminder, aes(continent)) +
  geom_bar() +
  ggtitle("Number of observations per continent")

```

### One qual + one quant variable
```{r boxplots, echo=TRUE}

# Filter out the data for 2007
gapminder2007 = gapminder %>% 
                  filter(year == 2007)

# Plot gdp per capita for 2007, all countries grouped by continent
ggplot(gapminder2007, aes(continent, gdpPercap)) + 
  stat_boxplot() +
  ggtitle("GDP per capita for each continent in 2007")

rm(gapminder2007) #clean up environment

```

Cool! We can see the plot of data from 2007 above, so let's work off of that and see if we can compare each year. 

```{r boxplots_allData, echo=TRUE}

length(unique(gapminder$year)) # there are 12 different years measured in these data

gapminder %>% 
  ggplot(aes(x=continent, y=gdpPercap)) +                   # pipe in data and aes
  stat_boxplot() +                                          # add boxplot
  facet_wrap(~ year) +                                      # plot in panels separated by year
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # arrange y-axis text vertically

```

Looks like the gdp per capita increases over the years and also becomes much more variable, especially in Asia and Europe. Interesting!

<a href="#top">Back to top</a>

## Use filter, select, and pipes! 

```{r gap_small, echo=TRUE}

# Plot the GDP of all countries (minus Oceania) in 1952 and 2007 
gapminder %>% 
  filter(continent != "Oceania", year == 1952 | year == 2007) %>%                      # filter out Oceania, keep only 1952 and 2007
  ggplot(aes(x=continent, y=gdpPercap)) +                                              # add ggplot aes
    stat_boxplot(alpha=1/2, size = 1, aes(color=continent), show.legend = FALSE) +     # color the boxes by continent plus adjust the transparency and size; don't show legend
    facet_wrap(~ year) +                                                               # separate panels by year
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +                         # adjust text to be vertical
    ggtitle("GDP per capita by continent (1952 vs. 2007)")                             # add title


# Add each individual country with "jitter"
# "jitter" helps spread the points out so it's easier to see! 
gapminder %>% 
  filter(continent != "Oceania", year == 1952 | year == 2007) %>% 
  ggplot(aes(x=continent, y=gdpPercap)) +
    stat_boxplot(alpha=1/2, size = 1, aes(color=continent), show.legend = FALSE) +
    geom_jitter() + 
    facet_wrap(~ year) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    ggtitle("GDP per capita by continent (1952 vs. 2007)")

```

```{r gap_asia}

asia = gapminder %>% 
  filter(continent == "Asia", year == 2007) %>%                 # filter only Asia in 2007
  select(country, lifeExp, gdpPercap, pop) %>%                  # remove year and continent
  arrange(gdpPercap) %>%                                        # arrange from smallest to largest GDP
  mutate(gdpGroup = factor(cut(gdpPercap, quantile(gdpPercap), include.lowest= TRUE), labels = c("low", "low-middle", "middle-high", "high"))) 
                                                                # split gdp into four quantiles

# plot life expectancy by gdp group
ggplot(asia, aes(gdpPercap, lifeExp)) +
  geom_point(alpha=3/4, size=3, aes(color=country)) +
  facet_wrap(~ gdpGroup) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Life expectancy of Asian countries in 2007, grouped by GDP per capita")

rm(asia) # cleanup environment

```

Let's explore some common factors in countries that lie below the life expectancy mean. We'll look at only data from 2007.

```{r lowLE, echo=TRUE}

# Create a new variable mutate
gap_mean = gapminder %>%
  mutate(LEgroup = ifelse(test = (gdpPercap < mean(gdpPercap)), yes = "low", no = "high")) %>%    #create new variable that keeps only the gdp's below the mean gdp
  filter(year == 2007) %>%                       # keep only data from 2007
  select(-year)                                  # get rid of the year variable (it's obsolete now)

head(gap_mean) # Countries now are classified as having "high" or "low" life expectancy

# histogram of GDP
gap_mean %>% 
  ggplot(aes(gdpPercap)) +
  geom_density() + 
  facet_wrap(~ LEgroup) +
  ggtitle("Distribution of countries by GDP per capita in 2007")

# histogram of population
gap_mean %>% 
  ggplot(aes(pop)) +
  geom_density() +
  facet_wrap(~ LEgroup) +
  ggtitle("Distribution of countries by population in 2007")

```

Hmmm... what are those outliers in the "low life expectancy" group? Let's check them out...

```{r remove_outliers, echo= TRUE}

# find the two outliers
justLOW = gap_mean %>% 
  filter(LEgroup != "high") %>%     # keep only rows that have "high LE"
  arrange(desc(pop))                # arrange them in descending order to see which ones are the highest outliers

# replot population histogram without outliers
gap_mean %>%
  filter(country != "China" & country != "India") %>%    # remove China and India
  ggplot(aes(pop)) +
  geom_density() +
  facet_wrap(~ LEgroup) +
  ggtitle("Distribution of countries by population (China and India removed) in 2007")

rm(justLOW) #clean up

```

Huh, without those outliers the histograms actually seem pretty similar! 

```{r continents_LEgroup, echo=TRUE}

# Barplot of countries in each continent, separated by LEgroup
gap_mean %>% 
  ggplot(aes(continent)) +
  geom_bar() +
  facet_wrap(~ LEgroup) +
  ggtitle("Number of countries within each continent in 2007, separated by low and high \nlife expectancies")

rm(gap_mean)

```

Check that out - there are no countries in Oceania that have a low life expectancy! Also, not surprisingly, the highest number of "low LE" countries are in Africa; the highest number of "high LE" countries are in Europe. 

<a href="#top">Back to top</a>

## More, more, more! 

```{r isThisRight, echo=TRUE}

# what's wrong with this? see below for answer!
new_gap = filter(gapminder, country == c("Rwanda", "Afghanistan"))

new_gap  

```

Oops! Nope, the analyst did *not* succeed in extracting all the data from Rwanda and Afghanistan. This function actually only extracted HALF of the data from each country? Why?? This was a little confusing at first, but after checking out the data, we can see that this filters the data in a pattern: first Rwanda, then Afghanistan, repeated over and over. That's because **country == c("Rwanda", "Afghanistan"))** tells filter function will go through that pattern as many times as possible until the end of the data. Here's the proper way to do it:

```{r}

# using the "or" argument
new_gap2 = filter(gapminder, country == "Rwanda" | country == "Afghanistan")

new_gap2

rm(new_gap, new_gap2)

```

Oh goody, we used the **or** argument. Now we have the full data from each country! Hooray!  

<a href="#top">Back to top</a>

## Reflection  
I liked this assignment! I think the hardest part was deciding what I wanted to plot, since there were so many variables that we could look at. Once I decided, using filter and select were pretty easy to integrate in. Dplyr is so user friendly, especially with pipes! Mutate was another function that seem easy enough to integrate.  

ggplot, on the other hand, is a bit more confusing for me. I feel like there are so many additions and moving pieces that it's sometimes hard to keep things straight. Simple plots are easy, but once you want to start grouping, adding color, creating multiple panels, etc., things get a little more complicated. Luckily, this [handy-dandy cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) exists, so that was really helpful!  

The other place I got stuck was trying to create a new variable with specific labels using the mutate function. I ended up using the ifelse function ("if-else"", not "i-felse" <-- I pronounced it like that for a good month before realizing what it really was...), but it took quite a bit of googling to figure it out. Handy little function, turns out! 

<a href="#top">Back to top</a>