---
title: "Class 10 notes"
author: "Kaitlyn Harper"
date: "October 5, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load}

library(tidyverse)
library(devtools)
install_github("JoeyBernhardt/singer")

library(singer)
data(locations)
data(songs)

View(locations)
View(songs)

```

## dplyr::join review

```{r}

left_join()
inner_join()
right_join()
full_join()
anti_join()
semi_join()

```

### Challenges

1. Produce a df with all the albums, artists, and the year in which the album was released 

```{r }

new_df = left_join(locations, songs, by="title") %>%  #add the songs data (containing year) to the locations df
  select(release, artist_name.x, year)  #keep only the variables you want

### This doesn't work:
new_df2 = left_join(songs, locations, by="title") %>%  
  select(release, artist_name.x, year) %>% 
  na.omit() #because one of the albums doesn't have a release YEAR - so it gets rid of it and you end up with only 13 rows instead of 14 

rm(new_df2)

```

2. Get the number of releases per year 

```{r}

# if you made a new df first (not dplyr)
table(new_df$year)

# if you want to pipe in directly (using dplyr functions)
inner_join(songs, locations, by="title") %>% 
  count(year)

rm(new_df, locations, songs)

```


## Reshaping

```{r load_large_df}

data("singer_locations")

```

**WIDE df vs LONG df**



```{r change_over_time}

small_df = singer_locations %>% 
  select(year, artist_hotttnesss, artist_familiarity, duration) %>% 
  filter(year != 0) #remove values that we don't know the year of 

small_df %>% 
  ggplot(aes(year, artist_familiarity))+
  geom_point()

```

Let's look at tidyr::gather

gather(df, key, value, set of columns)

```{r gather}

small_df_long = small_df %>% 
  gather(key = "Measure", value = "Units", artist_hotttnesss:duration)

small_df_long %>% 
  ggplot(aes(year, Units))+
  geom_point() +
  facet_wrap(~Measure, scales = "free_y") # SUPER HELPFUL CHAD TIP! 

```

```{r gather_again, echo=TRUE}

hfd_y = singer_locations %>% 
  select(year, duration:artist_familiarity) %>%  #### THIS IS IMPORTANT - SEE BELOW
  gather(key = "Measures", value = "My_value", 
         duration:artist_familiarity)

hfd_y %>% 
  filter(year > 1950) %>% 
  ggplot(aes(year, My_value))+
  geom_point() + 
  facet_wrap(~Measures, scales = "free_y") # SUPER HELPFUL CHAD TIP! 

# can also do "free" to free up both axes!! 
```

```{r spread}

hfd_spread = hfd_y %>% 
  spread(Measures, My_value)

```
^ Okay, so this doesn't work because you get a bunch of duplicates for 2007. IF you include the ID numbers (instead of selecting them out), spread will work.

```{r}

hfd_y = singer_locations %>% 
  select(year, duration:artist_familiarity) %>%  #### THIS IS THE STEP THAT MESSES YOU UP
  gather(key = "Measures", value = "My_value", 
         duration:artist_familiarity)

hfd_y_NEW = hfd_y = singer_locations %>% 
  gather(key = "Measures", value = "My_value", 
         duration:artist_familiarity)

### Note that we didn't select out any columns, so we have all the information still

hfd_y_NEW2 = singer_locations %>% 
  select(year, song_id, duration:artist_familiarity) %>%  ### This also works - just include the ID! 
  gather(key = "Measures", value = "My_value", 
         duration:artist_familiarity)

hfd_spread_NEW = hfd_y_NEW %>% 
  spread(Measures, My_value)

hfd_spread_NEW2 = hfd_y_NEW2 %>% 
  spread(Measures, My_value)

### voila! it works! 

rm(hfd_spread_NEW, hfd_spread_NEW2, hfd_y_NEW2, hfd_y)

```


### Reshape2 does about the same thing that tidyr does

```{r reshape2, echo=TRUE}

library(reshape2)

```

`dcast` is similar to `gather`
`melt` is similar to `spread`

```{r }

test = hfd_y_NEW %>% 
  dcast(year ~ Measures,        # number of rows, then which columns to spread (i.e. values in Measures)
        value.var = "My_value", # what values you use for the next argument
        fun.aggregate = var,    # this is what you're finding with the aggregate values from the argument above
                                  # IF there are multiple rows for a year
        na.rm=TRUE) %>%         # ignore NAs 
  na.omit()

```
With this function, we are creating one row for each value in the year column, and one column for each different value in the Measures column (duration, hotness, familiarity)





