---
title: 'Homework #3'
author: "Kaitlyn Harper"
date: "September 28, 2017"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data and packages
```{r load_data, echo=TRUE}

# Load packages
library(tidyverse)
library(knitr)

# Load gapminder dataset
library(gapminder)
data("gapminder")

#View data (just to make sure it's there)
head(gapminder)

```

<a href="#top">Back to top</a>

## Get the maximum and minimum of GDP per capita for all continents.

```{r max_min, results='asis'}

min_max = gapminder %>% 
  group_by(continent) %>% 
  summarise(maxGDP = max(gdpPercap), minGDP = min(gdpPercap))

knitr::kable(min_max, col.names = c("Continent", "Max GDP", "Min GDP"))

```

## Look at the spread of GDP per capita within the continents.

```{r spread, echo=TRUE}

gapminder %>%
  ggplot(aes(x=continent, y=gdpPercap)) +
  geom_boxplot(aes(fill = continent), show.legend = FALSE) +
  labs(x = "Country", y = "GDP per capita", title = "Spread of GDP per capita within each continent") +
  theme_dark(base_size = 13, base_family = "Courier")

```


## Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

## How is life expectancy changing over time on different continents?

## Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

## Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.
