---
title: 'Homework #3'
author: "Kaitlyn Harper"
date: "September 28, 2017"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data and packages
```{r load_data, message=FALSE}

# Load packages
library(tidyverse)
library(knitr)

# Load gapminder dataset
library(gapminder)
data("gapminder")

#View data (just to make sure it's there)
head(gapminder)

```

<a href="#top">Back to top</a>

### Look at the spread of GDP per capita within the continents.

```{r max_min, results='asis'}

min_max = gapminder %>% 
  group_by(continent) %>% 
  summarise(maxGDP = max(gdpPercap), 
            minGDP = min(gdpPercap),
            meanGDP = mean(gdpPercap),
            sdGDP = sd(gdpPercap))

knitr::kable(min_max, col.names = c("Continent", "Max GDP", "Min GDP", "Mean GDP", ""))

```

```{r spreadPlot, echo=TRUE}

gapminder %>%
  ggplot(aes(x=continent, y=gdpPercap)) +
  geom_boxplot(aes(fill = continent), show.legend = FALSE) +
  labs(x = "Country", y = "GDP per capita", title = "Spread of GDP per capita within each continent") +
  theme_dark(base_size = 13, base_family = "Courier")

```

<a href="#top">Back to top</a>

## How is life expectancy changing over time on different continents?  

```{r LEtable, echo=TRUE}

### make this table later

```


```{r LEplot, echo=TRUE}

gapminder %>%
  group_by(continent, year) %>% 
  summarise(meanLE = mean(lifeExp)) %>% 
  ggplot(aes(x=year, y=meanLE)) +
  geom_smooth(aes(color=continent), se=FALSE) +
  scale_color_discrete("Continents") +
  labs(x="Year", y= "Mean Life Expectancy (Years)", title="Life expectancy changes in each continent, 1952-2007") +
  theme_light(base_family = "Palatino") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_blank())

```

## Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

```{r lowLE, echo=TRUE}



```


## Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.
